Writing Trash Bin

### 2024-02-22 paper.qmd
moved and reworked a bunch of the context section

@@

## Perception Experiments and a Poor Definition
Even the most simple experiments are not free from the issues in the uncertainty definition.

Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results. These papers will be discussed in the upcomming sections.

### Contextual Inormation
I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty. I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point. I actually think this quote from the grammar of graphics is exactly what I was trying to say. *"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect. Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991). But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*. Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain. The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data. *This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*. This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. The confusion that is created in the literature was going to be the topic I had later. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction. Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.

The taxonomies that cannot be contained within the @utypo taxonomy are not a new uncertainty dimensions, but rather they are conflating adjacent information with uncertainty. A good example of this is the distinction between spatial, temporal, and estimated uncertainty. When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data. For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty. Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects. Once we have removed the physical nature these objects are near identical when considered statistically. Of course the nature of them can change based on what we choose to observe (i.e. if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g. whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline. It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it. The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty. It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data. While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.

In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to incorrect conclusions, we are merely suggesting for a different way of slicing the metaphysical cake of uncertainty that better aligns with what is *actually happening* when we perform a statistical analysis. This metaphysical cake is illustrated below as @fig-cake. In changing  what the actual boundaries are when we work with these problems, we improve our understanding of what we are actually doing when we conceptualise uncertainty, and in the process allow for more creative solutions creativity (e.g. by borrowing concepts from spatial visualisation to visualise uncertainty in photographs generated). However, since uncertainty is almost synonymous with the concept of "statistics", this adjustment in how we conceptualise and organise statistical methods would likely benefit other domains within the field.

![The metaphysical cake we are slicing differently](metaphysicalcake.jpg){#fig-cake}

- Specifies uncertainty can be represented by three components attribute (what) position (where) and temporal (when) @Kinkeldey2014
- Studies dealing with uncertainty information typically deal with attribute but rarely position and time @Kinkeldey2014
- 10/44 studies involve multiple types of uncertainty @Kinkeldey2014

- When we boil down data into a statistical object, we boil away many aspects on context. Birth and coin flips become an object that is one in the same, and often the visualisation 
- Why are spatial and temporal data considered special? What is the statistical element that remains after they have been away? Inexchangeability. Temporal data is one dimensional inexchangeability, spatial data has two dimensional inexchangeability.
  - There are other things that have inexchangeability, and not every question related to spatial or temporal data requires the inexchangeability.
- perceptual tasks need to know what you are uncertain about.
- Yes the data type is relevant but its not the most important thing
  - one with temporal, one with that temporal paper but you didn't know it was temporal, one with the bus time cdf
  - screenshot from these three papers, the connecting factor is the question, not the data type
  - The data type: spatial, temporal, etc, usually is mentioned as the 
  - Not a feature of the data, but it is a feature of the question "how do I show uncertainty on a map?". This is a fine question to ask, but the lack of specificity in what is being done, and the zoning in on functionally irrelevant elements leads the literature as a huge mess.
- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017

@@

@@

@@

>>>>

## 3.3 Contextual Information
Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty. These are things like spatial data, temporal data, category data, etc. While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

Papers that stress the importance of spatial or temporal contexts may be conflating contextual information with uncertainty. @fig-spatial depicts the role of contextual information in graphics, as something that helps us understand the information in the statistical graphic. When we create statistical models and quantify uncertainty, the spatial or temporal aspects are often boilled away [@Otsuka2023], and the only element of it that remains is the inexchangeability of our data. For this reason, there realistically is no such thing as "spatial" uncertainty in the same sense that there is no such thing as "photograph" uncertainty or "coin toss" uncertainty or "baby birth" uncertainty. Maps and photographs boil down to the same statistical objects in the same way that coin tosses and baby births boil down to the same statistical objects. Once we have removed the physical nature these objects are near identical when considered statistically. Of course the nature of them can change based on what we choose to observe (i.e. if we consider a coin landing on its side to be an outcome or if we consider intersex babies as an outcome in births) but this means the contextual information of our data, e.g. whether or not it is spatial or temporal, is only of importance at two stages of the data analysis pipeline. It is first important when we *boil it down* to our data through *observation* and again when we *map our statistical information back to its real world information* to *communicate* it. The only way the context uniquely contributes to uncertainty is in the "boil down" stage, however spatial or temporal uncertainty never refers to that specific cause of uncertainty. @Kinkeldey2014 discussed how uncertainty  can be represented by three components attribute (what) position (where) and temporal (when) and that studied typically deal with uncertainty around attribute but rarely position and time, however it is never specified what considerations should cause attribute uncertainty to be different to position or temporal uncertainty. @Kay2016 did an experiment that showed uncertainty around bus arrival times, however the visualisation used in the experiment, shown in @fig-bustime, is indistinguishable from most work that would be considered "attribute" uncertainty.

![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bustime}

It is typically in reference to measurement error, data aggregation, statistical modelling, or other steps that occur *after* the boil down stage, and are therefore not unique to spatial or temporal data. While the spatial or temporal features *may* become relevant at the final communication stage, however this has little do with quantifying uncertainty, and the difficulties in communicating a complicated statistical model through a complicated map is already well discussed in the visualisation literature, and in later sections we will discuss how this concept is related to uncertainty visualisation.

![Data pipeline with visualisations to show why we are ignoring contextual information for the time being. The visualisations that depict the data are similar to typical statistical graphics. The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}

This does not mean we are saying spatial and temporal information is never of importance. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. This position is best explained by @Leland2005 in *The Grammar of Graphics*:

> "Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break... but this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics."

This paper focuses on uncertainty, specifically uncertainty that is still in the the abstract *statistics* domain. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised inside the context provided by maps and, not what they should *try* to visualise based on the uncertainty abstraction. Additionally, since uncertainty is defined in the abstract domain, of the issues faced by spatial uncertainty visualisations are likely shared by other graphics that only differ in contextual information.

In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain. This is not only the case for uncertainty data. @Slingsby2023 discussed using a gridded glyphmap for spatial modelling, however the statistical elements are identical to a fluxuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011. The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background. @Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework. This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endevour, but these plots seem to be have been developed unaware of the existing statistical framework. When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-dicipline.

```{r}
#| output: asis
#| echo: false
img_files <- fs::dir_ls("prodplots", glob="*.png")
cat("::: {layout-ncol=2}\n",
    glue::glue("![]({img_files})\n\n\n"),
    ":::",
    sep = ""
)
```

@@

@@

## Perception Experiments, Question Problems, and Task Specificity
Uncertainty, like any other variable we are going to plot, has a natural mappings that should be considered when we express it in a plot. In the same way that mapping the colours red to hot and blue to cold, many variables, including uncertainty, have an intuitive mapping.

The choice of what we map uncertainty to will impact how easily

Even the most simple experiments are not free from the issues in the uncertainty definition.
Some experiments directly relate the internal 
Papers that consider unique aspects of uncertainty in their experiment such as experiemnts that identify the usefulness of specific perceptual tasks, experiments that identify problematic heuristics in commonly used uncertainty visualisations, and experiments that consider unique problems in uncertainty experiments all have useful results. These papers will be discussed in the upcomming sections.

-  Error is best mapped to fuzziness, location, and colour value; arrangement, size and transparency are an OK second choice; but saturation, hue, orientation and shape are unacceptable and have no intuitive connection to variance [@Maceachren2012]. No only do the graphical elements we map our features to matter, but the direction matters too. Graphical elements that are more fuzzy (fuzziness), further from centre (location), lighter (colour value), poorly arranged (arrangement), smaller (size), more transparent (transparency) are perceived to be more uncertain [@Maceachren2012]. This idea also extends to interval estimation, where questions about probability are best answered with gradients and questions about start and end times are easiest to answer with ambiguation [@Gschwandtnei2016]. Heuristics can work against us just as much as they can work for us. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]

- @Nathonours found that a scatter plot is better than a line plot if you want to convey the correlation between two time series but we cannot be sure if this was influenced by swapping the distribution depicted or by dropping the irrelevant feature of inexchangeability.

@@

The view of uncertainty as a dimensionality problem is not quite right. It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

- depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes. One colour represents the variable of interest while the other represents the sampling error of that variable. There are two immediate problems with this method. First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation. Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice. While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions. The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018]. Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018]. Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].

- The line-up protocol is an application on this idea and considers each visualisation to be a single outcome of some larger distribution [@Buja2009; @Wickham2010; @Chowdhury; @Hofmann2012]. By generating a sample of visualisations from a hypothetical distribution, visualisation authors can check if perceived patterns are real or merely the result of chance.

The final consideration when deciding what information to depict is whether or not additional information will clutter the graph. Our visualisation should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph]. It also seems that people will discount information they perceive as discountable if they are able. Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].

@@

## Bad Question stuff?
In the discussion of the HOPs, error bar, and violin plot study @Hullman2015 note:

> "In light of the common usage of error bars for presenting multiple independent distributions, it is noteworthy how poorly subjects using these representations performed on tasks asking them to assess how reliably a draw from one variable was larger than the other...Many subjects reported values less than 50%, which are not plausible given that the mean of B was larger than the mean of A."

- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016

### 2024-02-22 writingdumpster.qmd
moved and reworked a bunch of the context section

# Moved over notes
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016

>>>>

# Moved over notes
- mentions uncertainty can be variance, precision, accuracy, reliability, or related concepts @Hullman2016
-  @Spiegelhalter2017 defines risk as risk known probabilities and uncertainty as estimated probabilities, therefore uncertainty is what we have when you cannot accurately quantify risk.
- Authors provide little justification for their chosen response models (e.g. absolute accuracy vs relative measures) @Hullman2016 
- Some questions (e.g. what are the chances that the no.6 bus will arrive first) you can elicit the viewers subjective probability distribution @Hullman2016
- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- coloured maps for earthquake risk are easily interpretable @Spiegelhalter2017
- Hurricane risk is not only the path, but also the storm surge and wind speed. @Spiegelhalter2017
